Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-02-26 paper.tex
Paper edits

\cite{Gallagher1988EmpiricalModelPlasmasphere} modeled the the plasmasphere as a simple linear fit dependent on $L$ shell, geographic altitude, geomagnetic latitude, and $K_p$.  They found they could accurately model geomagnetically active conditions  \cite{Hamilton1988RingCurrentDevelopment} analyzed a strong geomagnetic storm and compared energy measurements of the ring current to $D_{st}$, along with measurements of ring current compositions before, during, and after the storm.  \inote{What did they find/conclude?}

\cite{Lotoaniu1999PlasmaMassDensity} used measurements from the \inote{???} instrument on the CRRES sateelite and compared in-situ measured plasma mass density to that inferred by ground-based ULF wave measurements that extended to the plasmatrough.  \inote{What did they find/conclude?}

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the SEM \inote{Check instrument} instruments on the GOES satellites from 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alf\v'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

@@

\section{Data Preparation}

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980 through 1991; all other parameters are from \cite{Reconstruction} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data is shown in Figure~\ref{AllData}.

In this work, events defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values.

Figure \ref{AllData} shows values of solar wind averages and mass density from GOES 6.  The corresponding 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day (non-overalpping?) windows using all non--fill hourly values.  A scatter plot of these two lines is shown in Figure~\ref{ccplot}.  The linear correlation is found to be $0.94$ matching the value found by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991.

@@

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-50$~nT.  The second is an increase in $\rho_{eq}$ above 40~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of the number of values used in computing the median divided by the square root of the number of values used.

In the upper panel of \ref{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $\sim 2$ amu/cm$^3$ instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of \ref{DailyAverages} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire run of GOES 6 over 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

@@

This method finds 669 such periods between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were (?) $D_{st}$ events and (?) $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of values used in computing the median dividied by the square root of the number of values used.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/alldata-GOES6-1983-1991.eps}
\caption{Overview of data used in this study. The top four panels show parameters from \cite{Reconstruction} and the bottom panel contains $\rho_{eq}$ based on GOES 6 measurements from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st} = -50$~nT and $\rho_{eq} = 40$~amu/cm$^3$}.
\label{AllData}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-day.eps}
\caption{Events in the interval 1989-1991. Upper panel: $D_{st}$ events. Lower panel: Events in the interval 1983-1991.}
\label{DailyAverages}
\end{figure}

>>>>

\cite{Gallagher1988EmpiricalModelPlasmasphere} used the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model the the plasmasphere as a simple linear fit dependent on $L$ shell, geographic altitude, geomagnetic latitude, and $K_p$.  They found they could accurately model geomagnetically active conditions by using a weighted average $K_p$ value and an azimuthal symmetry at smaller L-shells.  \cite{Hamilton1988RingCurrentDevelopment} analyzed a strong geomagnetic storm and compared energy measurements of the ring current to $D_{st}$, along with measurements of ring current compositions before, during, and after the storm. They found $D_{st}$ to be correlated with the inner ring current energy density during and after storm onset, an inward movement of energy density, and a rapid $D_{st}$ recovery from the quick loss of heavier ions.

\cite{Lotoaniu1999PlasmaMassDensity} used measurements from the Plasma Wave Experiment instrument on the CRRES satellite and compared in-situ measured plasma mass density to that inferred by ground-based ULF wave measurements from the CANOPUS array. By extending the ground footprints to the plasmatrough, they found agreement between the two methods for certain plasmapause conditions.  \vinote{Mention specifics? Worth including paper?}

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the GOES satellites from 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alf\v'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

@@

\section{Data Preparation and Overview}

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980 through 1991; all other parameters are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements \vinote{More explanation?}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in the respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 is found to be $0.94$ matching the value found by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991.  The corresponding 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of -50~nT.  The second is an increase in $\rho_{eq}$ above 40~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991 to correspond with the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter were computed using the standard deviation of the values used in computing the median divided by the square root of the number of values used.

In the upper panel of Figure \ref{DailyAverages}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $\sim 2$ amu/cm$^3$ instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events which is much larger than the number of \req\ values as we use all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed. \vinote{Alternate: Note that for $D_{st}$ events, we computed median values using all available data, even when \req\ availability was sparse.}

In the lower panel of Figure \ref{DailyAverages} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire run of GOES 6 over 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

@@

This method finds 669 periods of $D_{st}$ crossing beneath the selected threshold between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours \vinote{Make sure I defined "duration"}. Figure \ref{Storms} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{Storms}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that were used to compute the median. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack.

Figure \ref{Storms} shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $D_{st}$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

@@

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of values used in computing the median dividied by the square root of the number of values used.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.45]{paperfigures/alldata-GOES6-1983-1991.eps}
\caption{Overview of data used in this study. The top four panels show parameters from \cite{Kondrashov2014ReconstructionOfGaps} and the bottom panel contains $\rho_{eq}$ based on GOES 6 measurements from \cite{Denton} after interpolation and averaging described in the text. Dashed horizontal lines in the $D_{st}$ and $\rho_{eq}$ panels indicate sample event cutoff thresholds of $D_{st} = -50$~nT and $\rho_{eq} = 40$~amu/cm$^3$}.
\label{AllData}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-50-tak.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dst-day.eps}
\caption{$D_{st}$ events from GOES 6. Upper panel: Events in the interval 1989-1991. Lower panel: Events in the interval 1983-1991.}
\label{DailyAverages}
\end{figure}
